---
title: "Translation for package translators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Translation for package translators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(potools)
```

This vignette is aimed at package translators; if you're the developer of a package (or just want to understand how it looks from their site), you start with `vignette("developers")`.

First of all, thank you!
Providing translations for a package is a fantastic way to contribute and I know that your work is appreciated by the developer.
Next, check that the package has started the process of providing translations.
The easiest way to check is to see if there's a `po/` directory in the package source.
If not, you might want to start by opening an issue to see if the developer is interested in providing translations.

As a translator, you'll work primarily with `.po` files.
`.po` files start with some metadata, but mostly consist of pairs of `msgid` and `msgstring` that look like this:

    msgid "Original message"
    msgstr ""
