---
title: "Translation for package translators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Translation for package translators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(potools)
```

This vignette is aimed at package translators; if you're the developer of a package (or just want to understand how it looks from their site), you start with `vignette("developers")`.

First of all, thank you for helping out!
Providing translations for a package is a fantastic way to contribute and I know that your work is deeply appreciated by the package developer.

Before you go any further, make sure the that the package has started the process of providing translations by checking that it has `po/` directory in the package root.
If you don't see that package, you might want to start by opening an issue to see if the developer is interested in providing translations.

### Creating translations

Before we get into the details lets review the basic process of creating translations:

-   The package developer runs `po_extract()` to extract all translatable messages from their R and C code.

-   You call `po_create()` to generate a `.po` file for the language you're translating.
    A `.po` file consists of pairs of lines like:

        msgid "This is the message in English"
        msgstr ""

    You then replace each `msgstr` with the appropriate translation:

        msgid "This is the message in English"
        msgstr "This is the message in another language"

-   Finally, either you or the package developer uses `po_compile()` to turn the plain text `.po` files into binary `.mo` files that are distributed with the package.

The process for updating translations is slightly different, and we'll come back to it later in the doc.

## Translation basics

As a translator, you'll work primarily with `.po` files.
There's one `.po` file for each language that lives in the `po/` directory.
If you'll look at this directory, you'll also see a `.pot` file --- that's a template file that contains the list of all translatable strings in the package and is used to generate the `.po` files for each language.

Get started by creating a `.po` for your language by running `potools::po_create("{language code}")`.
Next, open that file with your favourite text editor.
At the top of the file you'll see some metadata that looks something like this:

    msgid ""
    msgstr ""
    "Project-Id-Version: potools 0.2.3\n"
    "Report-Msgid-Bugs-To: \n"
    "POT-Creation-Date: 2021-11-06 14:19-0700\n"
    "PO-Revision-Date: 2021-11-06 14:19-0700\n"
    "Last-Translator: Michael Chirico <michaelchirico4@gmail.com>\n"
    "Language-Team: ja\n"
    "Language: ja\n"
    "MIME-Version: 1.0\n"
    "Content-Type: text/plain; charset=UTF-8\n"
    "Content-Transfer-Encoding: 8bit\n"
    "X-Generator: potools 0.2.3\n"

Begin by updating `Last-Translator` to your email address; this tells the developer how to reach you (e.g. to ask if you might update your translations when a new version of the package comes out).

The rest of the file consist of pairs of `msgid` and `msgstr` that initially look like this:

    #: translate_package.R:66
    msgid "Running message diagnostics..."
    msgstr ""

The `msgid` is the string that appears in the R source code; it could be in an error, warning, message, or just printed to the console.
The `msgstr` is the translated equivalent and it's your job as a translator to fill out `msgstr` with the equivalent in your language:

    #: translate_package.R:66
    msgid "Running message diagnostics..."
    msgstr "メッセージ診断中。。。"

That's basically it!
Just work through the file filling in the translations one by one.
If there's not quite enough context to figure out the best translation you can either look at the source code (using the line file number reference in the line above `msgid`) or contact the developer for clarification.
(If a message is challenging to translate, it often indicates that the English version might be suboptimal, so most developers will appreciate you reaching out).

Next we'll go through the details of few message variations you might come across, show you how to try out your work, and then finish up by discussing a few extra details that arise when updating the translations for a package, rather than starting from scratch.

## Message variations

The following sections describe four important message variations:

-   Messages that use `glue()`
-   Messages that use `sprintf()`
-   Multi-line messages
-   Plurals

### `glue()`

You can tell if message uses [glue](https://glue.tidyverse.org) because it will contain a pair of braces: `{}`.
glue evaluating any R code in between `{}` and inserting the results into the string:

```{r}
name <- "Michael"
glue::glue("Hi {name}!")
```

The code above would generate the following `.po`:

    msgid: "Hi {name}!"
    msgstr: ""

The main thing to remember when translating glue strings is not to translate the contents of `{}` since that is the name of a variable in the code:

    msgid: "Hi {name}!"
    msgstr: "こんにちは {name}！"

### `sprintf()`

You can tell if a message uses `sprintf()` because it will contain a format specifier like `%s` (a string), `%i` (an integer), or `%f` (a floating point number).
When called, `sprintf()` replaces these placeholders with values:

```{r}
name <- "Michael"
sprintf("Hi %s!", name)
```

This would generate the following `.po`:

    msgid: "Hi %s!"
    msgstr: ""

Which (if you spoke Japanese), you might translate to:

    msgid: "Hi %s!"
    msgstr: "こんにちは %s！"

If there are multiple interpolated strings, and you need to change their order to make grammatical sense in your language, you can put `1$`, `2$` etc after the `%` to refer to variable in that position.

```{r}
sprintf("%s %s %s", "first", "second", "third")
sprintf("%2$s %1$s %3$s", "first", "second", "third")
```

This can be hard to puzzle out, which is why we recommend that package developers use `glue()`.

### Multi-line messages

If the message is very long, it might get wrapped across multiple lines:

    msgid ""
    "Reproducing these messages here for your reference since they might still "
    "provide some utility."
    msgstr ""

You don't need to worry about preserving the line breaks, since the translated version might be shorter or longer than the English version.
Just note that each line must start and end with `"` and that if it doesn't all fit on one line, the first `msgstr` should be `""`.

### Plurals

Messages that vary based on some count have slightly different form.
They begin with the English singular and plural forms:

    msgid "There is {n} cow in the field"
    msgid_plural "There are {n} cows in the field"

Then are followed by the forms for your language.
Depending on your language, you will need to provide somewhere between 1 and 6 translations.
For example, Japanese only needs a single form, because the rest of the sentence doesn't vary based on the number of cows:

    msgid "There is {n} cow in the field"
    msgid_plural "There are {n} cows in the field"
    msgstr[0]: フィールドには{n}頭の牛がいます

Slovenian has four forms (1, 2, 3-4, everything else) so gets four entries:

    msgid "There is {n} cow in the field"
    msgid_plural "There are {n} cows in the field"
    msgstr[0]: Na polju je {n} krava
    msgstr[1]: Na polju sta {n} kravi
    msgstr[2]: Na polju so {n} krave
    msgstr[3]: Na polju je {n} krav

(Note the forms are zero-indexed, so they start at 0, not 1.)

The rules for converting the number to the `msgid` index are expressed using C-code and can be found at the top of the `.po` file (look for "plural-forms").
It's a little tricky to go from the C code to a human description of the cases, but if you're a native speaker you're hopefully already familiar with the basic breakdown.

## Trying out your work

If you know how to use the package, you can try out your translations.
There are three important steps:

1.  Set the LANGUAGE environment variable to the language you're translating: `Sys.setenv(LANGUAGE "{language code"}`. (You only need to do this once per session).
2.  Compile the plain text `.po` file you've been editing to the binary `.mo` file that R uses by running `potools::po_compile()`.
3.  Re-load (with `devtools::load_all()`) or re-install the package (e.g. using RStudio's "Install and restart" button).

You can then work with the package as you usually would and any messages you've translated should appear in your language.

## Updating a package

So far we've assumed you're providing the first set of translations for a package.
But another common scenario is updating the translations for a new release of the package.
To get to this point, the developer will have updated the `.pot` file and all of the `.po` files.
There are three cases:

-   A message has been **added**: the `msgstr` is empty, and you'll need to fill it in like for the initial translation.

-   A message has been **removed**: old translations are moved to the bottom of the file (so you can still refer to them if they're needed), and marked as deprecated (using the ??? symbol).

-   A message has **changed**: if it's a big change, it'll be treated like an addition and a deletion, but if it's a small change it might get marked as fuzzy.

        #: build-news.R:285
        #, fuzzy
        msgid "CRAN release: %s"
        msgstr "Versión en CRAN %s"

    TODO: show how the old message shows up.

    You'll need to double check the translation and remove "fuzzy" once you're done.
